\chapter*{\sffamily Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

Firstly, I extend my gratitude to my supervisor, Andrés Marino Álvarez, and my co-supervisor, Germán Castellanos Domínguez. Their invaluable guidance and support have been indispensable in successfully completing this research. Their contributions have influenced this study and left an indelible impact on my academic journey.

My deepest acknowledgement is extended to my beloved wife who has consistently been a source of strength, providing unwavering support throughout every challenge. Additionally, I wish to express my heartfelt thanks to my dear parents and sister who have persistently stood by me. Their words of wisdom during challenging times and unconditional love have been my pillars of support and stability. Lastly, the most special acknowledgment is reserved for my son, Emanuel, who has been a beacon of joy and hope.

I am particularly grateful to all members of the Signal Processing and Recognition Group (SPRG) at the National University of Colombia in Manizales. A special note of thanks goes to Julian Caicedo and Mateo Tobón whose camaraderie and academic discussions have significantly enriched my experience over these years.

Finally, I recognize that this research would not have been possible without the support given by the projects: “Alianza científica con enfoque comunitario para mitigar brechas de atención y manejo de trastornos mentales y epilepsia en Colombia (ACEMATE).” (Code 111091991908, Hermes Code 56118 ) funded by MINCIENCIAS, and “ Sistema de visión artificial para el monitoreo y seguimiento de efectos analgésicos y anestésicos administrados vía neuroaxial epidural en población obstétrica durante labores de parto para el fortalecimiento de servicios de salud materna del Hospital Universitario de Caldas - SES HUC.” (Hermes Code 57661 ), funded by Universidad Nacional de Colombia.